Assessing Racial Bias in Arrest and Booking Practices in Toronto from 2020-2021

### 2024-12-04 paper/paper.qmd
Improving grammar

---
title: "Assessing Racial Bias in Arrest and Booking Practices in Toronto: Insights and Implications from 2020-2021"
subtitle: "Black and Indigenous Overrepresentation in Arrests with Stable Booking Probabilities Across Ethnic Groups"
author: "Akshat Aneja"
abstract: This study examines arrest and booking data from the Toronto Police Service for 2020 and 2021 to explore potential racial disparities in policing practices. While Black and Indigenous individuals were disproportionately represented in arrests, the likelihood of being formally booked after arrest was similar across ethnic groups. These findings suggest that disparities may occur earlier in the law enforcement process rather than at the booking stage. By highlighting these patterns, the study underscores the need for targeted reforms to ensure equity and transparency in policing practices.
date: today
date-format: long
number-sections: true
format: pdf
toc: true
thanks: "Code and data supporting this analysis is available at: https://github.com/Akshat211202/bias-in-arrests-toronto"

@@

This study examines disparities in booking probabilities across ethnic groups in Toronto during 2020 and 2021. While discussions about racial bias in policing often focus on arrest rates or use of force, booking decisions represent another critical stage in law enforcement processes that can reflect systemic bias.

Booking, the formal process of charging an individual after arrest, determines whether someone enters the criminal justice system. Examining booking probabilities provides insight into whether disparities exist in how different racial groups are treated after an arrest. This analysis focuses on identifying patterns in booking decisions for various ethnic groups and whether these patterns changed over the two years.

@@

The findings aim to provide objective evidence that can inform discussions about equity in policing. Identifying disparities in booking probabilities can help highlight areas where procedural changes may be necessary to ensure fair treatment for all groups. This study also explores how booking practices have evolved between 2020 and 2021 to identify potential progress or emerging trends.

@@

It is crucial to analyze the distribution of arrests across different age groups to understand potential patterns and disparities. [@tbl-table1] presents a summary of arrest counts for various ethnic groups across all age categories in 2020 and 2021. The age groups included in the dataset are: Under 17, 18 to 24 years, 25 to 34 years, 35 to 44 years, 45 to 54 years, 55 to 64 years, and 65 years and over.

Given the complexity and extensive nature of the dataset, which involves nine age groups across eight ethnicities, summarizing the entire dataset in a comprehensible manner is challenging. To address this and ensure clarity, we have selected a subset of the data focusing specifically on individuals aged 25 to 34 years. This group was chosen as it represents a significant proportion of the data and allows for a more straightforward comparison of arrest trends across ethnic groups while avoiding information overload. By narrowing the scope, the analysis becomes more manageable and the insights more focused.

@@

It is important to note that the OpenDataToronto [@rOpenDataToronto] provides population distribution data by ethnicity. However, due to the size and diversity of Toronto and its numerous wards, it has been challenging to cross-verify this data with crime rate statistics in a comprehensive manner. Further, a more detailed analysis could be conducted by comparing specific ethnic groups directly.

@@

Referring to [@tbl-table2], [@tbl-table3] and [@fig-fig1] the data reveals a non-uniform distribution of arrests across ethnic groups, with the Black and White populations showing the highest numbers of arrests in both 2020 and 2021. Additionally, the data indicates a decline in arrests for these two ethnic groups from 2020 to 2021. In contrast, arrests for other ethnic groups, such as South Asian, East Asian, and Indigenous populations, increased over the same period. These opposing trends highlight the divergent patterns of arrest across different ethnic groups during this time.

@@

The data reveals a non-uniform distribution of arrests across age groups, with 25 to 34 and 35 to 44 age brackets accounting for the highest number of arrests in both 2020 and 2021. However, there is a notable decrease in the number of arrests for these age groups of 2020 to 2021 which indicates that there may be a possible shift in law enforcement activity.

@@

The $\beta$ coefficients are important because they provide insight into how an individual’s race and the year of arrest influence the likelihood of being booked after arrest.

@@

The model was designed to provide insights into: Whether racial groups differ in booking probabilities after arrest. Whether booking probabilities changed between 2020 and 2021. The Bayesian framework ensures these insights are backed by a robust statistical foundation that accounts for both data-driven and prior uncertainties.

@@

Wide credibility intervals for racial group coefficients suggest that data limitations (e.g., small sample sizes for some groups) contributed to high variability in results. This highlights the need for more comprehensive data to draw robust conclusions.

@@

This report analyzed arrest and booking data for Toronto during 2020 and 2021 using a Bayesian logistic regression model to examine predictors of booking probabilities. The predictors in this analysis were ethnicity and year. The results provide insights into equity in booking practices and the implications of these findings for law enforcement policies.

@@

Analysis of arrest data revealed that Black and Indigenous individuals were overrepresented in arrests relative to their share of the population, consistent with prior findings from the Toronto Police Service and other studies on racial disparities in policing. However, this study focused on booking probabilities after arrest and found no significant differences in booking outcomes across ethnic groups after controlling for year.

@@

Communicating these findings is critical for informed public discourse. While arrest data often highlights disparities, the results suggest that booking probabilities may not exhibit the same biases. This distinction can help guide discussions on police reform and support targeted interventions to address inequities at all stages of law enforcement decision-making.

@@

The dataset relied on officer-recorded perceived race and binary booking outcomes, omitting variables such as offense type, search status, or socio-economic indicators. These omissions may obscure nuanced influences on booking decisions and contribute to the high uncertainty in coefficient estimates for smaller groups.

@@

1.  **Incorporate Contextual Variables**: Adding variables like offense type or search status would provide a more comprehensive view of factors influencing booking probabilities.
2.  **Examine Interaction Effects**: Analyzing interactions between race and year could highlight evolving trends for specific groups.
3.  **Use Hierarchical Models**: Regional analyses could identify precinct-level disparities or best practices in booking decisions.
4.  **Explore Arrest-to-Booking Pipelines**: Investigating disparities across the entire law enforcement process—from arrests to sentencing—would provide a holistic understanding of equity in policing.

>>>>

---
title: "Assessing Racial Bias in Arrest and Booking Practices in Toronto from 2020-2021"
subtitle: "Black and Indigenous Overrepresentation in Arrests with Stable Booking Probabilities Across Ethnic Groups"
author: "Akshat Aneja"
abstract: This study examines arrest and booking data from the Toronto Police Service for 2020 and 2021 to analyze potential racial disparities in policing practices. While Black and Indigenous individuals were disproportionately represented in arrests, the likelihood of being formally booked after arrest was similar across ethnic groups. These findings suggest that disparities may occur earlier in the law enforcement process rather than at the booking stage. By highlighting these patterns, the study underscores the need for targeted reforms to ensure equity and transparency in policing practices.
date: today
date-format: long
number-sections: true
format: pdf
toc: true
thanks: "Code and data supporting this analysis is available at: https://github.com/Akshat211202/bias-in-arrests-toronto"

@@

This study examines disparities in booking probabilities across ethnic groups in Toronto during 2020 and 2021. While discussions about racial bias in policing often focus on arrest rates or use of force, booking decisions represent another important stage in law enforcement processes that can reflect systemic bias.

Booking, the formal process of charging an individual after arrest, determines whether someone enters the criminal justice system. Examining booking probabilities provides understanding into whether disparities exist in how different racial groups are treated after an arrest. This analysis focuses on identifying patterns in booking decisions for various ethnic groups and whether these patterns changed over the two years.

@@

The findings aim to provide objective evidence that can inform discussions about equity in policing. Identifying disparities in booking probabilities can help highlight areas where procedural changes may be necessary to ensure fair treatment for all groups. This study also analyzes how booking practices have evolved between 2020 and 2021 to identify potential progress or emerging trends.

@@

It is important to analyze the distribution of arrests across different age groups to understand potential patterns and disparities. [@tbl-table1] presents a summary of arrest counts for various ethnic groups across all age categories in 2020 and 2021. The age groups included in the dataset are: Under 17, 18 to 24 years, 25 to 34 years, 35 to 44 years, 45 to 54 years, 55 to 64 years, and 65 years and over.

Given the complexity and extensive nature of the dataset, which involves nine age groups across eight ethnicities, summarizing the entire dataset in a thorough manner is challenging. To address this and ensure clarity, we have selected a subset of the data focusing specifically on individuals aged 25 to 34 years. This group was chosen as it represents a significant proportion of the data and allows for a more straightforward comparison of arrest trends across ethnic groups while avoiding information overload. By narrowing the scope, the analysis becomes more manageable and the findings more focused.

@@

It is important to note that the OpenDataToronto [@rOpenDataToronto] provides population distribution data by ethnicity. However, due to the size and diversity of Toronto and its numerous wards, it has been challenging to cross-verify this data with crime rate statistics in a thorough manner. Further, a more detailed analysis could be conducted by comparing specific ethnic groups directly.

@@

Referring to [@tbl-table2], [@tbl-table3] and [@fig-fig1] the data shows a non-uniform distribution of arrests across ethnic groups, with the Black and White populations showing the highest numbers of arrests in both 2020 and 2021. Additionally, the data indicates a decline in arrests for these two ethnic groups from 2020 to 2021. In contrast, arrests for other ethnic groups, such as South Asian, East Asian, and Indigenous populations, increased over the same period. These opposing trends highlight the divergent patterns of arrest across different ethnic groups during this time.

@@

The data shows a non-uniform distribution of arrests across age groups, with 25 to 34 and 35 to 44 age brackets accounting for the highest number of arrests in both 2020 and 2021. However, there is a notable decrease in the number of arrests for these age groups of 2020 to 2021 which indicates that there may be a possible shift in law enforcement activity.

@@

The $\beta$ coefficients are important because they provide understanding into how an individual’s race and the year of arrest influence the likelihood of being booked after arrest.

@@

The model was designed to provide understanding about whether racial groups differ in booking probabilities after arrest. Whether booking probabilities changed between 2020 and 2021. The Bayesian framework ensures these findings are backed by a robust statistical foundation that accounts for both evidence-based and prior uncertainties.

@@

Wide credibility intervals for racial group coefficients suggest that data limitations (e.g., small sample sizes for some groups) contributed to high variability in results. This highlights the need for more thorough data to draw robust conclusions.

@@

This report analyzed arrest and booking data for Toronto during 2020 and 2021 using a Bayesian logistic regression model to examine predictors of booking probabilities. The predictors in this analysis were ethnicity and year. The results provide a good understanding into equity in booking practices and the implications of these findings for law enforcement policies.

@@

Analysis of arrest data showed that Black and Indigenous individuals were overrepresented in arrests relative to their share of the population, consistent with prior findings from the Toronto Police Service and other studies on racial disparities in policing. However, this study focused on booking probabilities after arrest and found no significant differences in booking outcomes across ethnic groups after controlling for year.

@@

Communicating these findings is essential for informed public discourse. While arrest data often highlights disparities, the results suggest that booking probabilities may not exhibit the same biases. This distinction can help guide discussions on police reform and support targeted interventions to address inequities at all stages of law enforcement decision-making.

@@

The dataset relied on officer-recorded perceived race and binary booking outcomes, omitting variables such as offense type, search status, or socio-economic indicators. These omissions may obscure specific influences on booking decisions and contribute to the high uncertainty in coefficient estimates for smaller groups.

@@

1.  **Incorporate Contextual Variables**: Adding variables like offense type or search status would provide a more thorough view of factors influencing booking probabilities.
2.  **Examine Interaction Effects**: Analyzing interactions between race and year could highlight evolving trends for specific groups.
3.  **Use Hierarchical Models**: Regional analyses could identify precinct-level disparities or best practices in booking decisions.
4.  **Analyze Arrest-to-Booking Pipelines**: Investigating disparities across the entire law enforcement process—from arrests to sentencing—would provide a holistic understanding of equity in policing.
